---
layout: single
title: "Advent of Code, Day 2"
date:   2020-12-02
summary: "Tackling day two of Advent of Code"
type: rshiny
menu:
  sidebar:
    name: AdventofCode-02
    identifier: adventcode2020-2
    parent: rshiny
url: /adventcode-2020-02
hero: /images/rshiny/advent.jpg
---

Day two!

```{r setup, message = FALSE}
library(tidyverse)
library(janitor)
library(here)

knitr::opts_chunk$set(echo = TRUE)
```

### Problem

> To try to debug the problem, they have created a list (your puzzle input) of passwords (according to the corrupted database) and the corporate policy when that password was set.
>
> For example, suppose you have the following list:
>
> - 1-3 a: abcde
> - 1-3 b: cdefg
> - 2-9 c: ccccccccc
>
>Each line gives the password policy and then the password. The password policy indicates the lowest and highest number of times a given letter must appear for the password to be valid. For example, 1-3 a means that the password must contain a at least 1 time and at most 3 times.
>
>In the above example, 2 passwords are valid. The middle password, cdefg, is not; it contains no instances of b, but needs at least 1. The first and third passwords are valid: they contain one a or nine c, both within the limits of their respective policies.

### Input Data

```{r}
input_data <- read.delim("day-02.txt",header = FALSE)
```

### How many passwords are valid?

```{r}
password_table_one <- input_data %>% 
  separate(1,into = c("qty","character","password"),sep = " ") %>% 
  separate(qty, into = c("min","max"), sep = "-",convert = TRUE) %>% 
  mutate(character = str_remove(character,":"),
         password = str_split(password,""),
         count = map2_dbl(password,character,~sum(.x %in% .y)),
         valid = count <= max & count >=min)

head(password_table_one)

sum(password_table_one$valid)
```
