---
layout: single
title: "Advent of Code, Day 4"
date:   2020-12-04
summary: "Tackling day four of Advent of Code - passport security!"
type: rshiny
menu:
  sidebar:
    name: AdventofCode-04
    identifier: adventcode2020-4
    parent: rshiny
url: /adventcode-2020-04
hero: /images/rshiny/advent.jpg
---

Day four!

```{r setup, message = FALSE}
library(tidyverse)
library(here)

knitr::opts_chunk$set(echo = TRUE)
```

### Problem Description
> The automatic passport scanners are slow because they're having trouble detecting which passports have all required fields. The  expected fields are as follows:
> 
>     - byr (Birth Year)
>     - iyr (Issue Year)
>     - eyr (Expiration Year)
>     - hgt (Height)
>     - hcl (Hair Color)
>     - ecl (Eye Color)
>     - pid (Passport ID)
>     - cid (Country ID)
> 
> Passport data is validated in batch files (your puzzle input). Each passport is represented as a sequence of key:value pairs separated by spaces or newlines. Passports are separated by blank lines.

Count the number of valid passports - those that have `all required fields`. Treat cid as optional. In your batch file, how many passports are valid?

### Input Data

https://github.com/tanho63/advent_of_code/blob/master/2020/day-04.txt

```{r}
input_data <- read_lines("day-04.txt")

head(input_data)
```

Cleaning data into a tibble.

```{r}
df_passports <- tibble(data = input_data) %>%
  mutate(blank = ifelse(data == "",1,0),
         passport_id = cumsum(blank)+1) %>% 
  filter(!blank) %>% 
  select(-blank) %>% 
  separate_rows(data, sep = " ") %>% 
  separate(data, sep = ":", into = c("key","value")) %>% 
  pivot_wider(names_from = key, values_from = value)

head(df_passports)
```

### Problem 1: How many complete/non-missing passports?

```{r}

valid_p1 <- df_passports %>% 
  select(-cid) %>% 
  filter_all(~!is.na(.x))

head(valid_p1)
nrow(valid_p1)
```

### Problem 2: Validate columns

Time to add more validation beyond non-missing!

```{r}

valid_p2 <- valid_p1 %>% 
  mutate_at(c("byr","iyr","eyr"),as.numeric) %>% 
  filter(
    between(byr,1920,2002),
    between(iyr,2010,2020),
    between(eyr,2020,2030),
    case_when(
      str_ends(hgt,"cm") & between(parse_number(hgt),150,193) ~ TRUE,
      str_ends(hgt,"in") & between(parse_number(hgt),59,76) ~ TRUE,
      TRUE ~ FALSE
      ),
    str_detect(hcl,"^#[A-z,0-9]{6}$"),
    ecl %in% c("amb","blu","brn","gry","grn","hzl","oth"),
    str_detect(pid,"^[0-9]{9}$"),
  )

head(valid_p2)
nrow(valid_p2)

```
